---
title: "Data Preparation"
author: "Sarah Ryley"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r default load, include=FALSE}

rm(list=ls())
gc()

# Load default packages, report functions, disable scientific notation
source("./Functions/default.R")

```


# Prep federal data

```{r load dol data, include=FALSE}

#  22564969 rows 

uis <- read_csv("https://oui.doleta.gov/unemploy/csv/ar203.csv") %>%
  
  mutate() %>%

  # rename columns
  select(state = st,
         rptdate,
         sample_pop = c1,
         
         sex_male = c2,
         sex_female = c3,
         sex_ina = c4,
         
         eth_latinx = c40,
         eth_not_latinx = c41,
         eth_ina = c42,
         
         race_native_alaskan = c43,
         race_asian = c44,
         race_black = c45,
         race_hawaiian_pi = c46,
         race_white = c47,
         race_ina = c48,
         
         age_under_22 = c12,
         age_22_24 = c13,
         age_25_34 = c14,
         age_35_44 = c15,
         age_45_54 = c16,
         age_55_59 = c17,
         age_60_64 = c18,
         age_65_over = c19,
         age_ina = c20,
         
         industry_ag_forest_fish_hunt = c49,
         industry_mining = c50,
         industry_utilities = c51,
         industry_construction = c52,
         industry_manufacturing = c53,
         industry_wholesale_trade = c54,
         industry_retail_trade = c55,
         industry_transport_warehouse = c56,
         industry_information = c57,
         industry_finance_insurance = c58,
         industry_real_estate = c59,
         industry_prof_scientific_tech_services = c60,
         industry_management_of_cos_ent = c61,
         industry_admin_support_wastemgmt_remedial = c62,
         industry_educational_services = c63,
         industry_healthcare_social_assistance = c64,
         industry_arts_ent_rec = c65,
         industry_accomodation_food = c66,
         industry_other_services_excl_public_admin = c67,
         industry_public_admin = c68,
         industry_ina = c69,
         
         occ_mgmt = c70,
         occ_bussiness_financial_operations = c71,
         occ_computer_math = c72,
         occ_architecture_engineering = c73,
         occ_life_physical_social_sciences = c74,
         occ_communinity_social_services = c75,
         occ_legal = c76,
         occ_educational_library = c77,
         occ_arts_design_ent_sports_media = c78,
         occ_healthcare_practitioners_technical = c79,
         occ_healthcare_support = c80,
         occ_protective_sevice = c81,
         occ_food_prep_serving = c82,
         occ_building_grounds_cleaning_maintenance = c83,
         occ_personal_care_service = c84,
         occ_sales_related = c85,
         occ_office_admin_support = c86,
         occ_farming_fishing_forestry = c87,
         occ_construction_extraction = c88,
         occ_installation_maintenance_repair = c89,
         occ_production = c90,
         occ_transportation_material_moving = c91,
         occ_military = c92,
         occ_ina = c93)

```

```{r uis edit}

uis_edit <- uis %>%
  # calc total
  mutate(total = sex_male + sex_female + sex_ina, 
        
         # changed age category because many states used these combined categories in their initial claims data.
         age_24_under =  age_under_22 + age_22_24,
         age_55_64 = age_55_59 + age_60_64) %>%
  
  select(-c(age_under_22, age_22_24, age_55_59, age_60_64)) %>%
  select(state, rptdate, sample_pop, total, everything()) %>%
  
  gather("measure", "measure_n", sex_male:age_55_64) %>%
  separate(measure, c("measure_type", "measure"), sep = "_", extra = "merge") %>%
  
  # calc percent of total for measure using the information available
  group_by(state, rptdate, measure_type) %>%
  mutate(measure_ina = measure_n[which(measure == "ina")],
         measure_pct = case_when(measure != "ina" ~ measure_n/(total-measure_ina),
                                 TRUE ~ measure_n/total),
         measure_pct = replace_na(measure_pct, 0),
         measure_pct_ina = measure_ina/total) %>%
  ungroup() %>%
  
  # add variables
  left_join((read_csv("Input/state_names.csv")), by = "state") %>%
  mutate(claim_type = "Insured",
         # date variables
         rptdate = mdy(rptdate),
         year = year(rptdate),
         year_month = as.yearmon(rptdate),
         month = as.integer(month(rptdate)),
         
         # axis label
         month_labs = factor(month,
                             levels = 1:12,
                             labels = c("Jan","Feb","Mar","Apr","May","Jun",
                                        "Jul","Aug","Sep","Oct","Nov","Dec"))) %>%
  
  # so the median is for previous 25 years (1995-2019)
  filter(year >= 1995 & year <=2019) %>%
  
  # add measure stats
  group_by(state, measure_type, measure, month_labs) %>%
  mutate(measure_min_pct_month = min(measure_pct),
         measure_max_pct_month = max(measure_pct),
         measure_median_pct_month = median(measure_pct)) %>%
  
  as.data.frame()

make_report(uis_edit) %>%
  arrange(-`# Missing`)

#uis_edit %>% write_csv("Output/federal_unemployment_data_edit.csv")

```

<br>

#### Check federal edit

```{r uis checks}
 
# max date
uis_edit %>% summarise(max = as.yearmon(max(rptdate)),
                       min = as.yearmon(min(rptdate)))

uis_edit %>% group_by(year(rptdate)) %>% summarise(n=n_distinct(state))
# Doesn't hit 53 until 1998

# chart federal total receiving unemployment by month
uis_edit %>% group_by(year_month) %>% summarise(total = sum(total)) %>%
  ggplot(aes(x=year_month, y=total)) +
  geom_line()

# count number of months female share is 50% or greater.
uis_edit %>%
  filter(measure == "female") %>%
  count(measure_pct >= 0.5)

```

<br>
<br>

# Prep state data

```{r state}

states <- read_csv("Input/state_data_add.csv")

states_edit <- states %>%
  mutate_all(funs(replace_na(.,0))) %>%
  
  mutate(age_24_under = age_under_22 + age_22_24 + age_24_under,
         age_55_64 = age_55_64 + age_55_59 + age_60_64) %>%
  
  # standardize race columns to match US DOL race columns
  mutate(eth_latinx = 	eth_latinx +	racecombined_latinx +	
           raceeth_latinx_native +		
           raceeth_latinx_other +		
           raceeth_latinx_unknown +		
           raceeth_latinx_white ,		
         eth_not_latinx =	eth_not_latinx +	racecombined_native_alaskan +	racecombined_hawaiian_pi +
           raceeth_not_latinx_native +	racecombined_asian +	racecombined_white +
           raceeth_not_latinx_other +	racecombined_black +	racecombined_asian_pi +
           raceeth_not_latinx_unknown +		
           raceeth_not_latinx_white ,		
         eth_ina =	eth_ina +	racecombined_ina +	
           raceeth_unknown_native +		
           raceeth_unknown_other +		
           raceeth_ina +		
           raceeth_unknown_white,		
         race_native_alaskan = 	race_native_alaskan +	racecombined_native_alaskan +	
           raceeth_latinx_native +		
           raceeth_not_latinx_native +		
           raceeth_unknown_native,		
         race_asian =	race_asian +	racecombined_asian_pi +	
           racecombined_asian,	
         race_black = race_black + racecombined_black,	
         race_hawaiian_pi =	race_hawaiian_pi +	racecombined_hawaiian_pi,	
         race_white =	race_white +	raceeth_unknown_white +	
           raceeth_latinx_white +	racecombined_white +	
           raceeth_not_latinx_white,		
         race_ina =	race_ina +	racecombined_ina +	
           raceeth_latinx_unknown +		
           raceeth_not_latinx_unknown +		
           raceeth_ina) %>%
  
  # remove measures that won't join with federal data
  select(-c(age_under_22, age_22_24, age_55_59, age_60_64)) %>%
  select(-contains("raceeth_")) %>%
  select(-contains("racecombined_")) %>%

  mutate(total = sex_male + sex_female + sex_ina) %>%
  select(state, rptdate, total, everything()) %>%
  
  gather("measure", "measure_n", -c(state, rptdate, total)) %>%
  separate(measure, c("measure_type", "measure"), sep = "_", extra = "merge") %>%
  
  # calc percent of total, for non-inas using only what has values
  group_by(state, rptdate, measure_type) %>%
  mutate(measure_ina = measure_n[which(measure == "ina")],
         measure_pct = case_when(measure != "ina" ~ measure_n/(total-measure_ina),
                                  TRUE ~ measure_n/total),
         measure_pct = replace_na(measure_pct, 0),
         measure_pct_ina = measure_ina/total) %>%
  ungroup() %>%
  
  # add state name
  left_join((read_csv("Input/state_names.csv")), by = "state") %>%
  
  # add other variables
  mutate(claim_type = "Initial",
         rptdate = mdy(rptdate),
         year = year(rptdate),
         year_month = as.yearmon(rptdate),
         month = as.integer(month(rptdate)),
         
         # axis label
         month_labs = factor(month,
                             levels = 1:12,
                             labels = c("Jan","Feb","Mar","Apr","May","Jun",
                                        "Jul","Aug","Sep","Oct","Nov","Dec"))) %>%
  
  # add state's date of closures
  left_join((read_csv("Input/covid_dates.csv") %>%
               mutate_at(vars(starts_with("date")), funs(mdy(.))) %>%
               # calculate earliest closure date
               mutate(date_min_covid = pmin(date_school_closures, date_restaurant_closures, date_stay_in_place))), 
            by = "state_name") %>%
  
  # Code if week ending was before or after earliest shutdown date
  mutate(covid_color = ifelse(rptdate<date_min_covid, "Before shutdowns", "After shutdowns"),
         
         # variable indicates if state gave monthly or weekly data
         monthly_weekly_ind = ifelse(rptdate=="2020-03-31", "monthly", "weekly")) %>%
  
  group_by(state, covid_color) %>%
  
  # count of weeks after covid shutdowns, if state only gave one week, or gave monthly data, remove from timeline chart
  mutate(count_weeks = n_distinct(rptdate),
         state_filter = ifelse(count_weeks < 2 & covid_color == "After shutdowns", "remove", "keep")) %>%
  as_data_frame()

make_report(states_edit) %>% arrange(-`# Missing`)

states_edit %>% write_csv("Output/state_data_edit.csv")

```

<br>

## Create comparison chart

Compares the initial claims filed for the weeks after the first COVID-19 shutdowns with the min, median, and max for March of the previous years. Removes Alaska, which only provided one week of data, and Ohio and Oklahoma, which only provided monthly statistics.

```{r covid compare chart}

covid_compare <- states_edit %>%
  filter(covid_color == "After shutdowns" & state_filter != "remove") %>%
  
  # create new totals
  group_by(state) %>%
  mutate(total = sum(total[which(measure == "female")])) %>%

  group_by(state, measure_type) %>%
  mutate(measure_ina = sum(measure_n[which(measure == "ina")]),
         measure_pct_ina = measure_ina/total) %>%
  
  group_by(state, state_name, count_weeks, total, measure_type, measure, measure_ina) %>%

  summarise(measure_n = sum(measure_n),
            start_week = min(rptdate),
            end_week = max(rptdate)) %>%
  
  # filter(state == "OR" &	measure =="not_latinx") %>%
  mutate(measure_pct = case_when(measure != "ina" ~ (measure_n/(total - measure_ina)),
                                       TRUE ~ measure_n/total),
         measure_pct_ina = measure_ina/total) %>%
  
  # take only available measures
  filter(measure_n>0) %>%
  
  # add measure median pct march from federal data
  left_join(
    (uis_edit %>%
       filter(month == 3) %>%
       select(state, measure_type, measure, 
              measure_median_pct_march = measure_median_pct_month) %>%
       unique()),
    by = c("state", "measure_type", "measure")) %>%
  
  # calc percentage point diff
  mutate(measure_ppt_diff_after_shutdowns = 
           round(((measure_pct - measure_median_pct_march)*100),0)) %>%
  as.data.frame()

covid_compare %>% write_csv("Output/states_post_covid_initial_claims_compare_march_median.csv")

# function for readable table
compare_table <- function(table) {
  table %>%
  select(state = state_name,
         total_claims = total,
         group_type = measure_type,
         group = measure,
         total_group = measure_n,
         share_group = measure_pct,
         median_historical_share_group = measure_median_pct_march,
         ppt_difference = measure_ppt_diff_after_shutdowns,
         start_week,
         end_week,
         total_measure_missing_info = measure_ina,
         share_measure_missing_info = measure_pct_ina) %>%
dt_filter() %>%
  formatPercentage(c("Share Group", "Median Historical Share Group", "Share Measure Missing Info"),0)
}

covid_compare %>%
  filter(measure_n>0) %>%
  group_by(measure_type, measure) %>%
  summarise(n_states = n_distinct(state))
  

```

<br>
<br>

# Stat checks

<br>

Check stats for the story, ["Coronavirus Shutdowns Make Women the Majority of Unemployment Seekers in States Across U.S."](https://fullerproject.org/story/coronavirus-shutdowns-make-women-the-majority-of-unemployment-seekers-in-states-across-u-s/), by Sarah Ryley and Jessica Washington, The Fuller Project, published April 22, 2020.

<br>

> Over the previous 25 years, women have typically comprised just over one-third of all unemployment recipients in March.

```{r national}

# make table with federal totals
uis_edit %>%
  filter(measure == "female") %>%
  # create new total not including INA
  mutate(total = total-measure_ina) %>%
  group_by(year, month) %>%
  summarise(total = sum(total),
            female = sum(measure_n)) %>%
  mutate(share_female = female/total) %>%
  
  group_by(month) %>%
  summarise(min = min(share_female),
            median = median(share_female),
            max = max(share_female))

```

<br>

>Even during the Great Recession, women never topped more than 47 percent of recipients.

```{r recession}

uis_edit %>%
  filter(measure == "female" &
           rptdate >= "2007-12-01" & rptdate <= "2009-06-01") %>%
  # create new total not including INA
  mutate(total = total-measure_ina) %>%
  group_by(year, month) %>%
  summarise(total = sum(total),
            female = sum(measure_n)) %>%
  mutate(share_female = female/total) %>%
  arrange(-share_female)
  
```

<br>

>Combined, the state data obtained by The Fuller Project represents 4.5 million of the 20.1 million new claims filed between March 15 and April 11.

```{r overall}

states_edit %>%
  filter(measure == "female" & rptdate>="2020-03-21"
         # & rptdate < "2020-04-17"
           ) %>%
  mutate(total_ina = total-measure_ina) %>%
  group_by(rptdate) %>%
  summarise(total = sum(total),
            total_ina = sum(total_ina),
            female = sum(measure_n)) %>%
  mutate(share_female = female/total_ina) %>%
  adorn_totals()

```

<br>

>Seventeen states provided some or all of the information requested. In the states that provided weekly breakdowns, after governors issued the first social distancing orders, women’s share of all new unemployment claims surged to highs ranging from between 53 percent in Wyoming to 67 percent in Alabama.

Note: Oklahoma and Ohio provided monthly data. Alaska only provided one weeks' worth of data.

```{r 17 state summary}

states_edit %>%
  filter(covid_color == "After shutdowns" & measure == "female" 
       #  & rptdate != "2020-03-31"
         ) %>%
  group_by(state) %>%
  mutate(min_week = min(rptdate),
         max_share = max(measure_pct)) %>%
  filter(measure_pct == max_share) %>%
  select(state, max_share) %>%
  arrange(-max_share)
  
names(states_edit)

```

<br>

<br>

>In the 14 states that provided data broken down by week, women’s share of new claims after the first coronavirus shutdowns happened in mid-March was significantly higher than their median share of unemployment recipients that month over the previous 25 years — from 11 percentage points in Oregon to 33 percentage points in North Dakota. (In Oregon, nearly half the applications filed after the shutdowns were missing gender information.)

```{r female inc table}

covid_compare %>% 
  filter(measure == "female") %>%
  compare_table()

```

<br>

> In 12 of the 14 states that provided weekly data on age, these workers’ share of new claims was higher than the historic median by between 5 and 43 percentage points.

> In several states, the share of claimants who identified as black or Latino was lower, and the share who identified as white or Asian higher.

```{r covid_compare summary}

covid_compare %>%
  group_by(measure_type, measure) %>%
  summarise(count_states_increase_5ppt_after_shutdowns = n_distinct(state[which(measure_ppt_diff_after_shutdowns>=5)]),
            count_states_provided_measure = n_distinct(state),
            percent_states_increase_5ppt_after_shutdowns = 
              (n_distinct(state[which(measure_ppt_diff_after_shutdowns>=5)]))/
              (n_distinct(state)),
              
            min_ppt_increase_after_shutdowns = min(measure_ppt_diff_after_shutdowns),
            max_ppt_increase_after_shutdowns = max(measure_ppt_diff_after_shutdowns),
            min_share_claims = min(measure_pct),
            max_share_claims = max(measure_pct)) %>%
  filter(measure != "ina") %>%
  dt_filter() %>%
  formatPercentage(c("Percent States Increase 5ppt After Shutdowns", "Min Share Claims", "Max Share Claims"),0)

names(covid_compare)

```

<br>
<br>

# Graphics

<br>

## Weekly timeline chart

```{r weekky chart}

states_edit %>%
  filter(measure == "female" & state_filter != "remove") %>%
  left_join(
    (uis_edit %>%
       filter(measure == "female" & month == 3) %>%
       select(state, month, measure_median_pct_month) %>%
       unique()),
    by = c("state")) %>%
  group_by(state) %>%
  ggplot(aes(x=rptdate, y=measure_pct, color=covid_color)) +
  geom_hline(aes(yintercept=measure_median_pct_month)) +
  geom_line(aes(x=rptdate, y=measure_pct)) +
  geom_point(aes(fill=covid_color), size = 6) +
  scale_colour_manual(values = c("#e6384f", "#57c7c2")) +
  # scale_colour_manual(values = c("#57c7c2", "#e6384f")) +
  facet_wrap(~state_name, ncol=4) +
  theme_bw() +
  #theme_bw() +
  scale_y_continuous(limits = c(0.1, 0.9),
                     labels = scales::percent_format(accuracy = 1),
                     breaks = c(.25, .5, .75)) +
  labs(y="% New Unemployment Claims Female",
       title="Surge in Women's Share of New Unemployment Claims After COVID-19 Shutdowns",
       subtitle ="Circles represent women's share of initial claims. Black line represents women's median share of recipients in March, 1995-2019.",
       caption="The Fuller Project analysis of statistics provided by state agencies and U.S. Labor Department data, 1995-2019.") +
  theme(legend.position = "top",
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        plot.title = element_text(color = "black", face = "bold"),
        plot.caption = element_text(hjust = 1))

```

<br>

## Dumbell chart

```{r covid compare dots}

covid_compare %>%
  filter(measure == "female") %>% 
  select(state_name, 
         `% Initial claims female after shutdowns` = measure_pct, 
         `Median % Insured Female in March` = measure_median_pct_march) %>%
  arrange(state_name) %>%
  gather("measure", "value", `% Initial claims female after shutdowns`:`Median % Insured Female in March`) %>%
  ggplot(aes(y=state_name, x=value)) +
  geom_line(aes(x=value)) +
  geom_point(aes(y=state_name, x=value, color=measure), size = 6) +
  scale_colour_manual(values = c("#e6384f", "#57c7c2")) +
  theme_minimal() +
  scale_x_continuous(limits = c(0.1, 0.9),
                     labels = scales::percent_format(accuracy = 1),
                     breaks = c(.25, .5, .75)) +
  labs(title="Women became majority of new unemployment claimants after COVID-19 shutdowns",
       caption="The Fuller Project analysis of statistics provided by state agencies and U.S. Labor Department data, 1995-2019.") +
  theme(legend.position = "top",
        legend.title = element_blank(),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        plot.title = element_text(color = "black", face = "bold"),
        plot.caption = element_text(hjust = 1))

```